Apache Flink\texttrademark: Unified Stream and Batch Processing \\ in a Single Engine

### 2015-11-26 architecture.tex
Reflects the google doc. Further changes will be done in latex.
small changes

\section{System Architecture}

@@

\para{Flink's Runtime and APIS.} Flink's software stack can be seen in \autoref{fig:FlinkStack}. The core of Flink is the distributed dataflow engine, which executes dataflow programs. A Flink runtime program is a DAG of stateful operators connected with data streams. There are two core APIs in Flink: the DataSet API for processing finite data sets (often referred to as “batch processing”), and the DataStream API for processing potentially unbounded data streams (often referred to as “stream processing”). Flink’s core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create runtime programs executable by the engine. As such, it serves as the common fabric to abstract both bounded (batch) and unbounded (stream) processing. On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs (currently FlinkML, Gelly, and Table).

As depicted in \autoref{fig:process-model}, a Flink cluster comprises three types of processes: the client, the JobManager (JM - master), and one or more TaskManagers (TM - workers). The client takes the program code, transforms it to a dataflow graph, and submits that to the JobManager. This transformation phase also examines the data types (schema) of the data exchanged between operators and creates serializers and other type/schema specific code. DataSet programs additionally go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in \autoref{sec:batch}).

The JobManager coordinates the distributed execution of the dataflow. It tracks the state and progress of each operator and stream, schedules new operators, and coordinates checkpoints and recovery. In a high-availability setup, the JobManager persists a minimal set of metadata at each checkpoint to a fault-tolerant storage, such that a standby JobManager can reconstruct the checkpoint and recover the dataflow execution from there (see \autoref{sec:fault-tolerance}).

The actual data processing takes place in the TaskManagers. A TaskManager executes one or more operators that produce streams, and reports on their status to the JobManager. The TaskManagers maintain the buffer pools to buffer or materialize the streams, and the network connections to exchange the data streams between operators.

\begin{figure}
\centering
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=.88\textwidth]{figs/stack.pdf}
\label{fig:FlinkStack}
\caption{Flink's Software Stack.}
\end{minipage}
\begin{minipage}{.49\textwidth}
\centering
	\includegraphics[width=.95\textwidth]{figs/architecture.pdf}
    \caption{Flink's Process Model.}
    \label{fig:process-model}
\end{minipage}%
\end{figure}

>>>>

\section{System Architecture}
\label{sec:architecture}

@@

\para{Flink's Runtime and APIS.} Flink's software stack can be seen in \autoref{fig:stack}. The core of Flink is the distributed dataflow engine, which executes dataflow programs. A Flink runtime program is a DAG of stateful operators connected with data streams. There are two core APIs in Flink: the DataSet API for processing finite data sets (often referred to as “batch processing”), and the DataStream API for processing potentially unbounded data streams (often referred to as “stream processing”). Flink’s core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create runtime programs executable by the engine. As such, it serves as the common fabric to abstract both bounded (batch) and unbounded (stream) processing. On top of the core APIs, Flink bundles domain-specific libraries and APIs that generate DataSet and DataStream API programs (currently FlinkML (machine learning), Gelly (graph processing), and Table (SQL-like operations)).

As depicted in \autoref{fig:process-model}, a Flink cluster comprises three types of processes: the client, the Job Manager, and one or more Task Managers. The client takes the program code, transforms it to a dataflow graph, and submits that to the JobManager. This transformation phase also examines the data types (schema) of the data exchanged between operators and creates serializers and other type/schema specific code. DataSet programs additionally go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in \autoref{sec:batch}).

The JobManager coordinates the distributed execution of the dataflow. It tracks the state and progress of each operator and stream, schedules new operators, and coordinates checkpoints and recovery. In a high-availability setup, the JobManager persists a minimal set of metadata at each checkpoint to a fault-tolerant storage, such that a standby JobManager can reconstruct the checkpoint and recover the dataflow execution from there (see \autoref{sec:fault-tolerance}). The actual data processing takes place in the TaskManagers. A TaskManager executes one or more operators that produce streams, and reports on their status to the JobManager. The TaskManagers maintain the buffer pools to buffer or materialize the streams, and the network connections to exchange the data streams between operators.

\begin{figure}[t!]
\begin{minipage}{1.1\linewidth}
      \centering
      \hspace{-0.1\linewidth}
      \begin{minipage}{0.48\linewidth}
          \begin{figure}[H]
              \centering
			\includegraphics[width=.88\textwidth]{figs/stack}
			\caption{Flink's Software Stack.}
			\label{fig:stack}
          \end{figure}
      \end{minipage}
      % \hspace{0.05\linewidth}
      \begin{minipage}{0.48\linewidth}
          \begin{figure}[H]
				\includegraphics[width=.95\textwidth]{figs/architecture.pdf}
    			\caption{Flink's Process Model.}
    			\label{fig:process-model}
          \end{figure}
      \end{minipage}
  \end{minipage}
\end{figure}

### 2015-11-26 exec-model.tex
small changes

\noindent \TODO{Do we need the one below? We will need more context if we decide to keep this here. Maybe move to streams?}

\para{Ordering Guarantees.} Data exchange of one stream partition preserves the order of records. As a consequence, chains of operators that consume one stream partition maintain order of records. Operators receiving more than one stream partition merge the streams as records arrive, to minimize back pressure. As a consequence, the streaming dataflow gives no ordering guarantees for records after any form of repartitioning or broadcasting. The responsibility to deal with out-of-order records is delegated to the operator implementation. We found that this gives the most efficient design, as most operators do not require deterministic order, and operators that need to compensate for out-of-order arrivals (such as event time windows) can do that efficiently.

\para{Control Events.} Apart from exchanging data, streams in Flink communicate different types of control events. These are special events injected in the data stream by producing operators and are delivered in-order along with all other data records and events within a stream partition. The receiving operators react to these events by performing certain actions upon their arrival. Flink uses control records in multiple mechanisms, such as: \vspace{-2mm}
\begin{itemize}
\item Signaling the progress of event time within a stream partition (\autoref{sec:streaming-time}). \vspace{-3mm}
\item Coordinating checkpoints, dividing the stream into pre-checkpoint and post-checkpoint (\autoref{sec:fault-tolerance}). \vspace{-3mm}
\item Signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (\autoref{sec:batch-iterations}). \vspace{-3mm}
\end{itemize}

@@

Recovery from failures reverts all operator states to their respective states taken from the last successful snapshot, and restarts the input streams starting from the latest barrier. The maximum amount of re-computation needed upon recovery is limited to the amount of input records between two consecutive barriers. Furthermore,  we note that this strategy does not eliminate the possibility for partial recovery, when upstream backup is applied.
\vspace{1mm}
\noindent ABS provides several benefits:\vspace{-3mm}
\begin{itemize}
\item It guarantees exactly-once state updates without ever pausing the computation \vspace{-3mm}
\item It is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval. \vspace{-3mm}
\item It is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc, depending on the larger environment in which Flink is used.
\end{itemize}

>>>>

\para{Control Events.} Apart from exchanging data, streams in Flink communicate different types of control events. These are special events injected in the data stream by producing operators and are delivered in-order along with all other data records and events within a stream partition. The receiving operators react to these events by performing certain actions upon their arrival. Flink uses special types of control events such as: \vspace{-2mm}
\begin{itemize}
\item \textit{Checkpoint barriers} that coordinate checkpoints by dividing the stream into pre-checkpoint and post-checkpoint (\autoref{sec:fault-tolerance}). \vspace{-3mm}
\item \textit{Watermarks} signaling the progress of event time within a stream partition (\autoref{sec:streaming-time}). \vspace{-3mm}
\item \textit{Iteration barriers} signaling that a stream partition has reached the end of a superstep, in Bulk/Stale-Synchronous-Parallel iterative algorithms on top of cyclic dataflows (\autoref{sec:batch-iterations}). \vspace{-1mm}
\end{itemize}

Control events assume that a stream partition preserves the order of records. To this end, unary operators in Flink consuming a single stream partition, \emph{guarantee a FIFO order of records}. However, operators receiving more than one stream partition merge the streams as records arrive, in order to keep up with the streams' rate and avoid back pressure. As a result, streaming dataflows in Flink do not provide ordering guarantees after any form of repartitioning or broadcasting, and the responsibility of dealing with out-of-order records is left to the operator implementation. We found that this gives the most efficient design, as most operators do not require deterministic order (e.g., hash-joins, maps, etc.), and operators that need to compensate for out-of-order arrivals (such as event time windows) can do that more efficiently as part of the operator logic.

@@

Recovery from failures reverts all operator states to their respective states taken from the last successful snapshot, and restarts the input streams starting from the latest barrier. The maximum amount of re-computation needed upon recovery is limited to the amount of input records between two consecutive barriers. Furthermore,  we note that this strategy does not eliminate the possibility for partial recovery, when upstream backup is applied.

\vspace{1mm}
\noindent ABS provides several benefits:\vspace{-3mm}
\begin{itemize}
\item It guarantees exactly-once state updates without ever pausing the computation \vspace{-3mm}
\item It is completely decoupled other forms of control messages, e.g., by events that trigger the computation of windows, thus not restricting the windowing mechanism to multiples of the checkpoint interval. \vspace{-3mm}
\item It is completely decoupled from the mechanism used for reliable storage, allowing state to be backed up to file systems, databases, etc, depending on the larger environment in which Flink is used.
\end{itemize}

### 2015-11-26 streaming.tex
small changes

\subsection{Stateful Stream Processing}
While most operators in Flink’s DataStream API look like functional, side-effect free operators, they also support efficient stateful computations. State is critical to many applications such as machine learning model building, graph analysis, user session handling and window aggregations among others. There is a plethora of different types of states depending on the use case. For example, the state can be something as simple as a counter or a sum, or more complex, such as a classification tree or a large sparse matrix often used in machine learning applications. Stream windows can also be modelled, in principle, by stateful operators which assign records to continuously updated buckets kept in memory as part of an operator state.

In Apache Flink, state is made explicit, to cover all different use cases, and incorporated in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operation;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations. Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system allowing very flexible custom state management in the streaming applications. Flink’s checkpointing mechanism guarantees that any registered state is durably accessed with exactly-once update semantics.

@@

\subsection{Stream Windows}
Many incremental computations over unbounded streams are often evaluated over finite logical views, called windows. Apache Flink incorporates windowing within a stateful operator which is configured via a flexible declaration composed out of three core functions: a window assigner and optionally a trigger and an evictor. All three functions can be selected among a pool of common predefined implementations (e.g. sliding time window) or explicitly defined by the user (via UDFs).

More specifically, the assigner is responsible for assigning each record to logical windows. For example, this decision can be based on the timestamp of a record when it comes to event-time windows. Note that in the case of sliding windows, an element can belong to multiple logical windows. An optional trigger defines when the operation associated with the window definition is performed. Finally, an optional evictor determines which records to retain within each window. The window assignment process of Apache Flink is uniquely capable of covering all known window types such as periodic time/count-windows, punctuation, landmark, session and delta windows. Note that windows incorporate out-of-order processing seamlessly as in \cite{li2005semantics, akidau2015dataflow} and, in principle, subsumes these windowing models. For example, below is a window definition of 6 seconds that slides every 2 seconds (Assigner). The window results are computed once the watermark passes the end of the window (Trigger).

@@

Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines. This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.

>>>>

\subsection{Stateful Stream Processing}
While most operators in Flink’s DataStream API look like functional, side-effect free operators, they provide support for efficient stateful computations. State is critical to many applications such as machine learning model building, graph analysis, user session handling and window aggregations. There is a plethora of different types of states depending on the use case. For example, the state can be something as simple as a counter or a sum, or more complex, such as a classification tree or a large sparse matrix often used in machine learning applications. Stream windows can also be modelled, in principle, by stateful operators which assign records to continuously updated buckets kept in memory as part of the operator state.

To cover a lot of different use cases, in Flink state is made explicit and is incorporated in the API by providing: $i)$ Operator interfaces or annotations to statically register explicit local variables within the scope of an operator;  $ii)$ an OperatorState abstraction for marking partitioned key-value states and the associated operations. Users can also configure how the state is stored and checkpointed using the StateBackend abstractions provided by the system, allowing very flexible custom state management in the streaming applications. Flink’s checkpointing mechanism (\autoref{sec:fault-tolerance}) guarantees that any registered state is durable with exactly-once update semantics.

@@

\subsection{Stream Windows}
Incremental computations over unbounded streams are often evaluated over continously evolving logical views, called windows. Apache Flink incorporates windowing within a stateful operator which is configured via a flexible declaration composed out of three core functions: a window \textit{assigner} and optionally a \textit{trigger} and an \textit{evictor}. All three functions can be selected among a pool of common predefined implementations (e.g. sliding time windows) or can be explicitly defined by the user as UDFs.

More specifically, the assigner is responsible for assigning each record to logical windows. For example, this decision can be based on the timestamp of a record when it comes to event-time windows. Note that in the case of sliding windows, an element can belong to multiple logical windows. An optional trigger defines when the operation associated with the window definition is performed. Finally, an optional evictor determines which records to retain within each window. Flink's window assignment process is uniquely capable of covering all known window types such as periodic time/count-windows, punctuation, landmark, session and delta windows. Note that Flink's windowing capabilities incorporate out-of-order processing seamlessly as in \cite{li2005semantics, akidau2015dataflow} and, in principle, subsume these windowing models. For example, below is a window definition of 6 seconds that slides every 2 seconds (Assigner). The window results are computed once the watermark passes the end of the window (Trigger).

@@

Note that streams are already partitioned on a key before windowing, so the above is a \textit{local operator} that does not require coordination between machines. This mechanism can be used to implement a wide variety of windowing functionality \cite{akidau2015dataflow}.
